Coalition Formation Algorithm of Prosumers in a Smart Grid Environment

### 2015-02-09 Article.tex
alpha modif
typos

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
% Abstract
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{abstract}
In a smart grid environment, we study coalition formation of prosumers that aim at entering the energy market. It is paramount for the grid operation that the energy producers are able to sustain the grid demand in terms of stability and minimum production requirement. We design an algorithm that seeks to form coalitions that will meet both of these requirements:  a minimum energy level for the coalitions and a steady production level which leads to finding uncorrelated sources of energy  to form a coalition. We propose an algorithm that uses graph tools such as correlation graphs or clique percolation to form coalitions that meet such complex constraints. We validate the algorithm against a random procedure and show that, it not only performs better in term of social welfare for the power grid, but also that it is more robust against unforeseen production variations due to changing weather conditions for instance. 
\end{abstract}

@@

There seems to be a clear consensus on the benefit of having bidirectional communication flows between the prosumers and the grid, as demand side management relies on such an architecture \cite{WuHamedHuangBook2011}. Nevertheless as the number of active prosumers is expected to rise, it is safe to assume there is a need for more complex communication patterns (prosumers grouping together into so-called "coalitions") that will help to decrease the communication burden and satisfy the multiple requirements of the power grid management.

Formation of coalitions inside a smart grid environment could be applied to various types of agents such as self sufficient micro-grids \cite{Pahwa}, sizable and adjustable virtual power plants (VPP) \cite{Braun, Ramchurn}, or fleet of electric vehicles that back up the grid in emergency situations (V2G) \cite{Ramchurn}. These are only a few use cases where the coalition of multiple agents can enhance the grid reliability and in the mean time reduce the communication burden.

In this paper, we consider the case where prosumers coalesce into virtual groups (the so-called coalitions) in order to be able to produce enough power which may be sold to the grid. If the coalition's production is sufficiently high and stable, the coalition can negociate on a day by day basis with the grid by announcing each day an expected production for the next day.

We focus on how to statistically improve the production stability by carefully forming coalitions of prosumers that have greater probabilities of staying in acceptable range of production. By using prediction techniques, we form coalitions that meet the contract values ranges proposed by the grid, enabling it to schedule its production accordingly. Meeting contract values for the prosumers is especially relevant in power trade conditions, where energy is traded based on day-ahead predictions.

@@

In order to select groups of prosumers that lower the volatility of the coalition's productions, we define an algorithm based on a popular approach for stocks market clustering \cite{Mantegna1999}. We use a modified clique percolation algorithm (see section \ref{sec:forming}) that will enable us to expand the cliques as needed in order to form proper coalitions that fulfill both the grid requirements and lower the production volatility. Finally, section \ref{sec:results} provides some results and a conclusion in section \ref{sec:conclusion}.

@@

Traditionally, forming coalitions in a pool of agents can be done either in a centralized way where a single central unit is responsible for all the computations or in a distributed way where agents have only local knowledges and take actions accordingly. It is of common use to represent the situation and assess the stability of the solution by using game theoretic tools. Some papers \cite{Saad2009} \cite{Luan2014} focus then on finding an optimal coalition structure giving a pool of autonomous self interested agents using distributed merge/split algorithms.

@@

$\epsilon$-graphs, consists in filtering edges based on their weight, only keeping edges whose weights are less than $ \epsilon $. In \cite{Garas2008, Onnela2004}, the authors studied the topological properties (average clustering, connectivity, relative number of cliques) of the correlation graph against those of growing random graphs, depending on the threshold $ \epsilon $. However, there is no well defined method to select a right tradeoff $ \epsilon $ as a function of the network topology.

@@

However, we will see in the following, that a clique (a complete subgraph) in our filtered decoralation-based graph exhibit, by construction, low correlation. We indeed removed in the filtering (see $\epsilon$-graphs) all links that stands for high correlation. There exists greedy heuristics based on cliques in the litterature such as clique percolation \cite{Lancichinetti} which realizes local optimizations of a fitness function and results in overlaping community structures.

Detection of overlapping communities is actually a very active field of research, especially in social networks where a person might belong to several communities. Section \ref{sec:forming} will explain how we used such techniques on the decorelation graph in order to form coalitions of prosumers.

@@

Our major concern is to characterize a prosumer behavior with realistic patterns of consumption, production as well as realistic geographical correlations between them. Only wind turbines and PV (photovoltaic panels) will be used as generators throughout this paper, and we denote by $ \mathcal{A} = \{ a_{1},...a_{N} \} $ the set of prosumers. To account for the production of a prosumer, we rely on French's weather data from 2006 to 2012 sampled every 3 hours \cite{Infoclimat} (similar data are available for the United States covering 2010\cite{NCDC}).

As shown in the first part of figure 1, the first step consists in discretizing the studied zone around well chosen weather stations and gathering three kinds of traces (wind speed, cloudiness, and temperature) that we organize in a structure we will refer to as "climate vector" in the following. We consider that climate vectors are constant over their area, meaning that, if two agents are in the same area, they are exposed to the same climate vector. We use existing DER models enabling conversion between weather variables and output powers (see power curves in \cite{Kota2011} \cite{windturbinemodel}). \cite{Dans2007} provides also a convenient way of using cloudiness traces as realistic degradation factors in a \textit{"clear blue sky"} solar model, which enables us to rebuild realistic solar irradiance traces for stations that do not provide such information directly.

@@

\begin{itemize}
\item \textbf{Daily cycles} : Consumption is low during night, and higher during the day with two picks in the morning and evening. Some noise is added so that prosumers have similar but not identical cycles.
\item \textbf{Seasonal cycles} : Consumption is higher in the winter because of heating and low in the summer (air conditioning is not considered). Temperature traces are used for modeling these cycles.
\end{itemize}

@@

Let $ P_{i}(t) $ be the instantaneous available power of agent \textit{i} at instant \textit{t} (his instantaneous production minus his instantaneous consumption). During the simulation (from $t_{0} $ to $ t_{K} $), all agents record their values with an hour time interval. As expected, the prosumer's time-series exhibit high seasonal patterns, but completely different from agent to agent because they depend both on the energy mix and the habits. Nevertheless, these macro variations can hide the interesting information in the correlation coefficients and twist the rest of our process. Thereby, we remove these seasonal effects for all agents (see second block of figure \ref{Fig1}). We denote by $ \mathcal{T}_{i} = \{ P_{i}(t_{0}),...,P_{i}(t_{K}) \} $ the resulting time-series for  agent \textit{i}, and by $ \mathcal{F}(P_{i}) $ the corresponding probability distribution.

@@

Let us recall that our goal is to group prosumers into coalitions so that the global power production resulting of the superposition of individual prosumer's productions be sufficiently predictible and above a minimum power level. We thus extend the agent notations to any coalition $ s \subset \mathcal{A} $ : 
\begin{itemize}
\item $ P_{s}(t) = \sum_{i \in s} P_{i}(t) $
\item $ \mathcal{T}_{s} = \{ P_{s}(t_{0}),...,P_{s}(t_{K}) \} $ with $ \mathcal{F}(P_{s}) $ the corresponding probability distribution
\end{itemize}

@@

\begin{table}[t]
\centering
\scriptsize
\begin{tabular}{l|p{4.5cm}}
  \hline 
  $ d_{ij} $ & Distance metric between timeseries i and j \\
  $ \rho_{ij} $ & Pearson's correlation coefficient between timeseries i and j \\
  $ \mathcal{A} $ & Agent set \\
  $ P_{i}(t) \in \mathbb{R} $ & Instantaneous production of agent $i$ at time $t$ \\
  $ P_{s}(t) $ & Instantaneous production of coalition s \\
  $ \mathcal{F}(P_{i}) $ & Probability distribution of $i$'s production \\
  $ \mathcal{F}(P_{S}) $ & Probability distribution of coalition s production \\
  $ \mathcal{N}(\mu, \sigma) $ & Normal distribution with mean $ \mu $ and standard deviation $ \sigma $ \\
  $ \phi \in [0,1] $ & Reliability threshold \\
  $ P^{MIN} \in \mathbb{R}^+ $ & Min value threshold \\
  $ P_{s}^{CRCT} \in \mathbb{R}^+ $ & Contract value announced by coalition s \\
  $ P_{\phi}(s) \in \mathbb{R}^+ $ & Max contract value under $ \phi $ constraint \\
  $ \mathcal{U}_{\phi,\ P^{MIN}}(s) $ & Utility function \\
  $ \tau_{i}(s) \in \mathbb{R}$ & Overlapping to disjoint coalitions mapping \\
  $ \epsilon \in [0,1] $ & Pruning threshold for the correlation graph \\
  $ G_{\epsilon}(\mathcal{A}, E) $ & Filtered correlation graph \\
  $ N_{COAL} \in \mathbb{Z}^+ $ & Number of desired coalition \\
  $ CS $ & Set of all coalition structures \\
  $ \Theta(G) $ & Set of non overlaping cliques in graph G \\
  $ \mathbf{1}_{X} $ & Indicator function over set X \\ 
  \hline
\end{tabular}

@@

\subsection{Utility function}\label{subsec:UtilityFunc}
As solving the optimization problem (see eq. \ref{problem}) is considered NP-hard, we address this problem with a heuristic based on a utility function (eq. \ref{utility}) that derives directly from the problem constraints. That is, if a coalition cannot provide a valid contract value, it receives a utility of zero. Furthermore, the utility increases with the overall production (sum of the contracted values) as stated in the objective of eq.\ref{problem}. The stability of the coalitions is also present in this utility function through the $ \phi $ parameter. Finally, the $ 1/|s| $ term in eq. \ref{utility} indicates that we favor small coalitions, mainly because they are easier to maintain.

\begin{equation}
 \mathcal{U}_{\phi,\ P^{MIN}}(s) = \mathbf{1}_{\textit{s\ valid}} \dfrac{P_{\phi}(s)}{|s|} 
\label{utility}
\end{equation}

@@

The model we used in order to simulate time-series of available production provides some diversity because of the combination between the energy mix (renewable generators combination) and climate vectors. Nevertheless, as the number of agents grows, the time-series tend to exhibit similar patterns. This is apparent when creating a correlation graph with similar metric such as defined in \cite{Garas2008} or \cite{Onnela2004} ($ 1 - \rho_{ij}^{2} $), where well defined clusters appear in the $ \epsilon $-graph for any values of $ \epsilon $.

@@

However, as seems intuitively understandable, cliques of this graph tend to exhibit very good utility values. Such  structures contain indeed a link between every two nodes, meaning that the overall correlation is quite small. Obviously, the $ \epsilon $ parameter is indirectly responsible for the sizes of the cliques : if it is too low, $ G_{\epsilon} $ does not provide enough cliques, conversely, if $\epsilon $ is too high, we loose important information as the graph becomes very dense and cliques tend to overlap strongly. For large values of $ \epsilon $ and for non trivial number of agents, finding cliques can even become computationally intensive.

Despite being direct and simple, improving the sizes of the cliques by increasing $ \epsilon $ does not seem elegant. Furthermore, as the utility function is also focused on maximizing $ P_{\phi}(s) $, it might be the case that some cliques benefit from additional agents even if they don't form a clique anymore. Cliques for small values of $ \epsilon $ appear thus as good seeds for forming stable coalitions rather than directly acceptable coalitions. We define $ N_{COAL} $ as the desired number of coalitions fixed by the user. Furthermore, let $ \Theta(G_{\epsilon}) $ be the set of non overlaping cliques in $ G_{\epsilon} $, we choose the $ \epsilon $ parameter as the smallest value such that $ G_{\epsilon} $ contains at least $ N_{COAL} $ non overlaping cliques :

@@

As explained in section \ref{sec:related}, clique percolation leads generally to overlapping communities. For simplicity, in this paper, we wish to keep the coalitions disjoint and  leave the management of overlapping coalitions for future work. We thus implemented a simple heuristic that considers nodes in multiple seeds one by one and chooses its final coalition as the one that “needs it the most” in terms of utility loss. More formally, for a coalition \textit{s} and a node $ i \in s $, we define\footnote{\label{note1}Not to be mingled with the Shapley value concept} :

@@

As shown in figure \ref{Fig2}a, the parameters $\phi$ and $P^{MIN}$ shape the utility function such that, if $ \phi $ is close to zero, the reliability requirement is very high. Only small values of $ P^{MIN}$ could then potentially lead to valid coalitions (and positive utilities). Conversely, the higher $\phi$, the less constraints are imposed to the coalitions and more valid coalitions could be formed for a larger spectrum of $ P^{MIN}$. Obviously, the highest utility values are found for high $ \phi $, because the formed coalitions are able to announce higher contract values, yielding higher utilities.

@@

In fig. \ref{Fig3}, we perform a simulation with 200 prosumers, and we fix the reliability $ \phi = 0.1 $ meaning that coalitions should produce more than their contract values at least $ 90 \% $ of the time. We observe the result of our algorithm for a fixed number of coalitions of $N_{COAL} = 10$ and let the algorithm form the coalitions as the grid changes according to $ P^{MIN} $. We compare our algorithm with a random process that splits the agents in $ N_{COAL} $ coalitions. Fig. \ref{Fig3} shows this evolution for our algorithm (blue dots) and for the random process (red dots). The diameter of a dot is proportional to the number of agents in the coalition and the higher the dot, the higher its utility. The $ P^{MIN} $ values of the x axis are expressed in tenth of MW for readability and the hatched zone corresponds to the \textit{"under-requirement space"}, meaning that whenever a coalition is in this zone, it has a null utility. Looking at fig. \ref{Fig3}, we can see first that our algorithm performs better than the random formation algorithm at finding high valued coalitions. The blue dots allowed to enter the market (non hatched zone) are indeed outnumbering the red dots, especially when the grid requirements are neither too low nor too high ($ P^{MIN} = 5 $ in fig. \ref{Fig3}).

@@

As visible on figure \ref{Fig5}, the production densities at the coalition level exhibit very different shapes. While our algorithm results in a nice density with a low (compared to the other curves) standard deviation, the other methods seem to form coalitions that will have a very large spectrum for their production. In more details, figure \ref{Fig4}a presents the evolution of social welfare as the number of coalitions increases (all other parameters are kept constant) for the same four methods. As for figure \ref{Fig4}b, it shows the percentage of coalitions able to enter the market for different values of $ P^{MIN} $. For consistency, we average the results of both plots over 100 realizations (colored zones around the curves shows the standard deviations). When the grid requirements are constant (fig \ref{Fig4}b), and the number of desired coalitions is low, clique percolation performs better than random case. Moreover, when $ N_{COAL} $ gets bigger, the performance of a random split tumbles down rapidly while our clique percolation is able to maintain efficiently the social welfare of the coalitions formed.

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
% Abstract
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{abstract}
In a smart grid environment, we study the coalition formation of prosumers that aim at entering the energy market. It is paramount for the grid operation that the energy producers are able to sustain the grid demand in terms of stability and a minimum production requirement. We design an algorithm that seeks to form coalitions that will meet both of these requirements:  a minimum energy level for the coalitions and a steady production level, which leads to finding uncorrelated sources of energy  to form a coalition. We propose an algorithm that uses graph tools such as correlation graphs or clique percolation to form coalitions that meet such complex constraints. We validate the algorithm against a random procedure and show that, it not only performs better in terms of social welfare for the power grid, but also that it is more robust against unforeseen production variations due to changing weather conditions for instance. 
\end{abstract}

@@

There seems to be a clear consensus on the benefit of having bidirectional communication flows between the prosumers and the grid, as demand side management relies on such an architecture \cite{WuHamedHuangBook2011}. Nevertheless, as the number of active prosumers is expected to rise, it is safe to assume there is a need for more complex communication patterns (prosumers grouping together into so-called "coalitions") that will help to decrease the communication burden and satisfy the multiple requirements of the power grid management.

The formation of coalitions inside a smart grid environment could be applied to various types of agents such as self-sufficient micro-grids \cite{Pahwa}, sizable and adjustable virtual power plants (VPP) \cite{Braun, Ramchurn}, or fleet of electric vehicles that back up the grid in emergency situations (V2G) \cite{Ramchurn}. These are only a few use cases where the coalition of multiple agents can enhance the grid reliability and in the meantime reduce the communication burden.

In this paper, we consider the case where prosumers coalesce into virtual groups (so-called coalitions) in order to be able to produce enough power which may be sold to the grid. If the coalition's production is sufficiently high and stable, the coalition can negociate on a day by day basis with the grid by announcing each day an expected production for the next day.

We focus on how to statistically improve the production stability by carefully forming coalitions of prosumers that have greater probabilities of staying in acceptable ranges of production. By using prediction techniques, we form coalitions that meet the contract values ranges proposed by the grid, enabling it to schedule its production accordingly. Meeting contract values for the prosumers is especially relevant in power trade conditions, where energy is traded based on day-ahead predictions.

@@

In order to select groups of prosumers that lower the volatility of the coalition's productions, we define an algorithm based on a popular approach for stock market clustering \cite{Mantegna1999}. We use a modified clique percolation algorithm (see section \ref{sec:forming}) that will enable us to expand the cliques as needed in order to form proper coalitions that fulfill both the grid requirements and lower the production volatility. Finally, section \ref{sec:results} provides some results and a conclusion in section \ref{sec:conclusion}.

@@

Traditionally, forming coalitions in a pool of agents can be done either in a centralized way where a single central unit is responsible for all the computations or in a distributed way where agents have only local knowledge and take actions accordingly. It is of common use to represent the situation and assess the stability of the solution by using game theoretic tools. Some papers \cite{Saad2009} \cite{Luan2014} focus then on finding an optimal coalition structure giving a pool of autonomous self-interested agents using distributed merge/split algorithms.

@@

$\epsilon$-graphs, consists in filtering edges based on their weight, only keeping edges whose weights are less than $ \epsilon $. In \cite{Garas2008, Onnela2004}, the authors studied the topological properties (average clustering, connectivity, relative number of cliques) of the correlation graph against those of growing random graphs, depending on the threshold $ \epsilon $. However, there is no well-defined method to select a right tradeoff $ \epsilon $ as a function of the network topology.

@@

However, we will see in the following, that a clique (a complete subgraph) in our filtered decorelation-based graph exhibit, by construction, low correlation. We indeed removed in the filtering (see $\epsilon$-graphs) all links that stand for high correlation. There exists greedy heuristics based on cliques in the litterature such as clique percolation, \cite{Lancichinetti} which realizes local optimizations of a fitness function and results in overlapping community structures.

Detection of overlapping communities is actually a very active field of research, especially in social networks where a person might belong to several communities. Section \ref{sec:forming} will explain how we used such techniques on the decorrelation graph in order to form coalitions of prosumers.

@@

Our main concern is to characterize a prosumer behavior with realistic patterns of consumption, production as well as realistic geographical correlations between them. Only wind turbines and PV (photovoltaic panels) will be used as generators throughout this paper, and we denote by $ \mathcal{A} = \{ a_{1},...a_{N} \} $ the set of prosumers. To account for the production of a prosumer, we rely on French's weather data from 2006 to 2012 sampled every 3 hours \cite{Infoclimat} (similar data are available for the United States covering 2010\cite{NCDC}).

As shown in the first part of figure 1, the first step consists in discretizing the studied zone around well-chosen weather stations and gathering three kinds of traces (wind speed, cloudiness, and temperature) that we organize in a structure we will refer to as "climate vector" in the following. We consider that climate vectors are constant over their area, meaning that, if two agents are in the same area, they are exposed to the same climate vector. We use existing DER models enabling conversion between weather variables and output powers (see power curves in \cite{Kota2011} \cite{windturbinemodel}). \cite{Dans2007} provides also a convenient way of using cloudiness traces as realistic degradation factors in a \textit{"clear blue sky"} solar model, which enables us to rebuild realistic solar irradiance traces for stations that do not provide such information directly.

@@

\begin{itemize}
\item \textbf{Daily cycles} : Consumption is low during the night, and higher during the day with two picks in the morning and evening. Some noise is added so that prosumers have similar but not identical cycles.
\item \textbf{Seasonal cycles} : Consumption is higher in the winter because of heating and low in the summer (air conditioning is not considered). Temperature traces are used for modeling these cycles.
\end{itemize}

@@

Let $ P_{i}(t) $ be the instantaneous available power of agent \textit{i} at instant \textit{t} (his instantaneous production minus his instantaneous consumption). During the simulation (from $t_{0} $ to $ t_{K} $), all agents record their values with an hour time interval. As expected, the prosumer's time-series exhibit high seasonal patterns, but completely different from agent to agent because they depend both on the energy mix and their habits. Nevertheless, these macro variations can hide the interesting information in the correlation coefficients and twist the rest of our process. Thereby, we remove these seasonal effects for all agents (see block 2 of figure \ref{Fig1}). We denote by $ \mathcal{T}_{i} = \{ P_{i}(t_{0}),...,P_{i}(t_{K}) \} $ the resulting time-series for  agent \textit{i}, and by $ \mathcal{F}(P_{i}) $ the corresponding probability distribution.

@@

Let us recall that our goal is to group prosumers into coalitions so that the global power production resulting of the superposition of individual prosumer's productions be sufficiently predictable and above a minimum power level. We thus extend the agent notations to any coalition $ s \subset \mathcal{A} $ : 
\begin{itemize}
\item $ P_{s}(t) = \sum_{i \in s} P_{i}(t) $
\item $ \mathcal{T}_{s} = \{ P_{s}(t_{0}),...,P_{s}(t_{K}) \} $ with $ \mathcal{F}(P_{s}) $ the corresponding probability distribution
\end{itemize}

@@

\begin{table}[t]
\centering
\scriptsize
\begin{tabular}{l|p{4.5cm}}
  \hline 
  $ d_{ij} $ & Distance metric between timeseries i and j \\
  $ \rho_{ij} $ & Pearson's correlation coefficient between time series i and j \\
  $ \mathcal{A} $ & Agent set \\
  $ P_{i}(t) \in \mathbb{R} $ & Instantaneous production of agent $i$ at time $t$ \\
  $ P_{s}(t) $ & Instantaneous production of coalition s \\
  $ \mathcal{F}(P_{i}) $ & Probability distribution of $i$'s production \\
  $ \mathcal{F}(P_{S}) $ & Probability distribution of coalition s production \\
  $ \mathcal{N}(\mu, \sigma) $ & Normal distribution with mean $ \mu $ and standard deviation $ \sigma $ \\
  $ \phi \in [0,1] $ & Reliability threshold \\
  $ P^{MIN} \in \mathbb{R}^+ $ & Min value threshold \\
  $ P_{s}^{CRCT} \in \mathbb{R}^+ $ & Contract value announced by coalition s \\
  $ P_{\phi}(s) \in \mathbb{R}^+ $ & Max contract value under $ \phi $ constraint \\
  $ \mathcal{U}_{\phi,\ P^{MIN}}(s) $ & Utility function \\
  $ \tau_{i}(s) \in \mathbb{R}$ & Overlapping to disjoint coalitions mapping \\
  $ \epsilon \in [0,1] $ & Pruning threshold for the correlation graph \\
  $ G_{\epsilon}(\mathcal{A}, E) $ & Filtered correlation graph \\
  $ N_{COAL} \in \mathbb{Z}^+ $ & Number of coalitions \\
  $ CS $ & Set of all coalition structures \\
  $ \Theta(G) $ & Set of non overlapping cliques in graph G \\
  $ \mathbf{1}_{X} $ & Indicator function over set X \\ 
  \hline
\end{tabular}

@@

\subsection{Utility function}\label{subsec:UtilityFunc}
As solving the optimization problem (see eq. \ref{problem}) is considered NP-hard, we address this problem with a heuristic based on a utility function (eq. \ref{utility}) that derives directly from the problem constraints. That is, if a coalition cannot provide a valid contract value, it receives a utility of zero. Furthermore, the utility increases with the overall production (sum of the contracted values) as stated in the objective of eq.\ref{problem}. The stability of the coalitions is also present in this utility function through the $ \phi $ parameter. Finally, the $ 1/|s|^{\alpha} $ term in eq. \ref{utility} controls the sizes of the coalitions (for simplicity, in the rest of the paper $ \alpha $ is set to one).

\begin{equation}
 \mathcal{U}_{\phi,\ P^{MIN}}(s) = \mathbf{1}_{\textit{s\ valid}} \dfrac{P_{\phi}(s)}{|s|^{\alpha}} 
\label{utility}
\end{equation}

@@

The model we used in order to simulate time-series of available production provides some diversity because of the combination between the energy mix (renewable generators combination) and climate vectors. Nevertheless, as the number of agents grows, the time-series tend to exhibit similar patterns. This is apparent when creating a correlation graph with similar metric such as defined in \cite{Garas2008} or \cite{Onnela2004} ($ 1 - \rho_{ij}^{2} $), where well-defined clusters appear in the $ \epsilon $-graph for any values of $ \epsilon $.

@@

However, as seems intuitively understandable, cliques of this graph tend to exhibit very good utility values. Such  structures contain indeed a link between every two nodes, meaning that the overall correlation is quite small. Obviously, the $ \epsilon $ parameter is indirectly responsible for the sizes of the cliques : if it is too low, $ G_{\epsilon} $ does not provide enough cliques, conversely, if $\epsilon $ is too high, we lose important information as the graph becomes very dense and cliques tend to overlap strongly. For large values of $ \epsilon $ and for non trivial number of agents, finding cliques can even become computationally intensive.

Despite being direct and simple, improving the sizes of the cliques by increasing $ \epsilon $ does not seem elegant. Furthermore, as the utility function is also focused on maximizing $ P_{\phi}(s) $, it might be the case that some cliques benefit from additional agents, even if they don't form a clique anymore. Cliques for small values of $ \epsilon $ appear thus as good seeds for forming stable coalitions rather than directly acceptable coalitions. We define $ N_{COAL} $ as the desired number of coalitions fixed by the user. Furthermore, let $ \Theta(G_{\epsilon}) $ be the set of non overlapping cliques in $ G_{\epsilon} $, we choose the $ \epsilon $ parameter as the smallest value such that $ G_{\epsilon} $ contains at least $ N_{COAL} $ non overlapping cliques :

@@

As explained in section \ref{sec:related}, clique percolation leads generally to overlapping communities. For simplicity, in this paper, we wish to keep the coalitions disjoint and  leave the management of overlapping coalitions for future work. We thus implemented a simple heuristic that considers nodes in multiple seeds one by one and chooses its final coalition as the one that “needs it the most” in terms of utility loss. More formally, for a coalition \textit{s} and a node $ i \in s $, we define :

@@

As shown in figure \ref{Fig2}a, the parameters $\phi$ and $P^{MIN}$ shape the utility function such that, if $ \phi $ is close to zero, the reliability requirement is very high. Only small values of $ P^{MIN}$ could then potentially lead to valid coalitions (and positive utilities). Conversely, the higher $\phi$, the less constraints are imposed to the coalitions and more valid coalitions could be formed for a larger spectrum of $ P^{MIN}$. Obviously, the highest utility values are found for high $ \phi $, because formed coalitions are able to announce higher contract values, yielding higher utilities.

@@

In fig. \ref{Fig3}, we perform a simulation with 200 prosumers, and we fix the reliability $ \phi = 0.1 $ meaning that coalitions should produce more than their contract values at least $ 90 \% $ of the time. We observe the result of our algorithm for a fixed number of coalitions of $N_{COAL} = 10$ and let the algorithm form the coalitions as the grid changes according to $ P^{MIN} $. We compare our algorithm with a random process that splits the agents in $ N_{COAL} $ coalitions. Fig. \ref{Fig3} shows this evolution for our algorithm (blue dots) and for the random process (red dots). The diameter of a dot is proportional to the number of agents in the coalition and the higher the dot, the higher its utility. The $ P^{MIN} $ values of the x axis are expressed in tenth of MW for readability and the hatched zone corresponds to the \textit{"under-requirement space"}, meaning that whenever a coalition is in this zone, it has a null utility. Looking at fig. \ref{Fig3}, we can see first that our algorithm performs better than the random formation algorithm at finding high-valued coalitions. The blue dots allowed to enter the market (non hatched zone) are indeed outnumbering the red dots, especially when the grid requirements are neither too low nor too high ($ P^{MIN} = 5 $ in fig. \ref{Fig3}).

@@

As visible on figure \ref{Fig5}, the production densities at the coalition level exhibit very different shapes. While our algorithm results in a nice density with a low (compared to the other curves) standard deviation, the other methods seem to form coalitions that will have a very large spectrum for their production. In more details, figure \ref{Fig4}a presents the evolution of social welfare as the number of coalitions increases (all other parameters are kept constant) for the same four methods. As for figure \ref{Fig4}b, it shows the percentage of coalitions able to enter the market for different values of $ P^{MIN} $. For consistency, we average the results of both plots over 100 realizations (colored zones around the curves show the standard deviations). When the grid requirements are constant (fig \ref{Fig4}b), and the number of desired coalitions is low, clique percolation performs better than random case. Moreover, when $ N_{COAL} $ gets bigger, the performance of a random split tumbles down rapidly while our clique percolation is able to maintain efficiently the social welfare of the coalitions formed.
